---
title: "PDT2 - EDA ANLY525"
author: "Ral Mogbogu"
date: "2025-05-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
getwd()
```


```{r library}
library(ggplot2) # Visualization
library(dplyr) # Data manipulation
library(caret)  # Predictive modeling 
library(corrplot) 
library(tidyr) # Data transformation
library(reshape2)
library(readxl)
```
```{r}
library(randomForest)
```
```{r}
install.packages("gbm")
library(gbm)
```
```{r}
library(MASS)
```


```{r}
diabetes.df <- read_excel("~/ANLY 525 Project/diabetes_health_indicators.xlsx")

```


```{r}
str(diabetes.df)
```

```{r smaller sample size}
set.seed(123)
diab.df <- diabetes.df[sample(nrow(diabetes.df), 3000), ]

View(diab.df)
```


```{r}
str(diab.df)
```


```{r}
library(skimr)
skim(diab.df)
```


```{r}
apply(diab.df,2,function(x) sum(is.na(x)))
```


```{r}
summary(diab.df)
```

```{r outliers}
#Boxplots
boxplot(diab.df$BMI, main = "Outliers in BMI", ylab = "BMI")
boxplot(diab.df$MentHlth, main = "Outliers in Mental Health Distribution", ylab = "MentHlth")
boxplot(diab.df$PhysHlth, main = "Outliers in Physical Health Distribution", ylab = "PhysHlth")


```


```{r check oultier columns}
diab.df[diab.df$BMI > 60, ]
diab.df[diab.df$MentHlth > 30, ] #no outlier cause scale is from 1-30 days
diab.df[diab.df$PhysHlth > 30, ] #no outlier cause scale is from 1-30 days
```


```{r}

ggplot(diab.df, aes(y = BMI)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Boxplot of BMI",
  y = "Body Mass Index (BMI)") +
  theme_minimal()

library(ggplot2)

ggplot(diab.df, aes(y = MentHlth)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Mental Health",
       y = "MentHlth") +
  theme_minimal()

library(ggplot2)

ggplot(diab.df, aes(y = PhysHlth)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.colour = "red", outlier.shape = 16) +
  labs(title = "Boxplot of Physical Health",
       y = "Physical Health (PhysHlth)") +
  theme_minimal()
```

```{r copy}
copy_diab.df = diab.df
str(copy_diab.df)

```

```{r}
unique(copy_diab.df$DiabOutcome) 

unique(copy_diab.df$Sex)

unique(copy_diab.df$GenHlth)
unique(copy_diab.df$Income)
unique(copy_diab.df$Age)

unique(copy_diab.df$HighBP)
unique(copy_diab.df$HighChol)
unique(copy_diab.df$CholCheck)
unique(copy_diab.df$Smoker)
unique(copy_diab.df$HeartDiseaseorAttack)
unique(copy_diab.df$PhysActivity)
unique(copy_diab.df$Veggies)
unique(copy_diab.df$HvyAlcoholConsump)
unique(copy_diab.df$AnyHealthcare)
unique(copy_diab.df$NoDocbcCost)
```


```{r}
# Should return: 0 1 2

# Step 2: Convert 0 and 1 (No Diabetes, Prediabetes) → 0
# And 2 (Diabetes) → 1
copy_diab.df$DiabOutcome <- ifelse(copy_diab.df$DiabOutcome == 2, 1, 0)

# Step 3: Turn into a factor for modeling
copy_diab.df$DiabOutcome <- factor(copy_diab.df$DiabOutcome,
                                       levels = c(0, 1),
                                       labels = c("No Diabetes", "Diabetes"))

# Step 4: Check final result
table(copy_diab.df$DiabOutcome)

# re-coding no diabetes AND prediabetes as 0, and diabetes as 1
```

```{r}


# Set up a layout for multiple plots (adjust nrow and ncol as needed)
numeric_cols <- ncol(copy_diab.df)
par(mfrow = c(ceiling(sqrt(numeric_cols)), floor(sqrt(numeric_cols))))
par(mar = c(4, 4, 2, 1), oma = c(0, 0, 2, 0)) # Adjust margins

# Create histograms for each column
for (col in names(diab.df)) {
  hist(diab.df[[col]],
       breaks = 20, # Equivalent to bins=20
       main = col,
       xlab = col,
       ylab = "Frequency")
}

title("Histograms of Variables in Diabetes Indicator Dataset", outer = TRUE) # Equivalent to plt.suptitle
```

```{r}
unique(copy_diab.df$DiabOutcome) 

unique(copy_diab.df$Sex)

unique(copy_diab.df$GenHlth)
unique(copy_diab.df$Income)
unique(copy_diab.df$Age)

unique(copy_diab.df$HighBP)
unique(copy_diab.df$HighChol)
unique(copy_diab.df$CholCheck)
unique(copy_diab.df$Smoker)
unique(copy_diab.df$HeartDiseaseorAttack)
unique(copy_diab.df$PhysActivity)
unique(copy_diab.df$Veggies)
unique(copy_diab.df$HvyAlcoholConsump)
unique(copy_diab.df$AnyHealthcare)
unique(copy_diab.df$NoDocbcCost)
```


```{r factor}

copy_diab.df$Sex <- factor(copy_diab.df$Sex, 
                           levels = c(0, 1), labels = c("Female", "Male"))

copy_diab.df$GenHlth <- factor(copy_diab.df$GenHlth, 
                               levels = 1:5, 
                               labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"), ordered = TRUE)

copy_diab.df$Income <- factor(copy_diab.df$Income, 
                              levels = 1:8, 
                              labels = c("<10k", "10–15k", "15–20k", "20–25k", "<35k", "<50k", "<75k", "75k+"), ordered = TRUE)

copy_diab.df$Age <- factor(copy_diab.df$Age, 
                           levels = 1:13, 
                           labels = c("18–24", "25–29", "30–34", "35–39", "40–44","45–49", "50–54", "55–59", "60–64", "65–69", "70–74", "75–79", "80+"), ordered = TRUE)

```


```{r levels defined}
# 0 is No, 1 is Yes
copy_diab.df$HighBP <- factor(copy_diab.df$HighBP, 
                              levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$HighChol <- factor(copy_diab.df$HighChol, 
                                levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$CholCheck <- factor(copy_diab.df$CholCheck, 
                                 levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$Smoker <- factor(copy_diab.df$Smoker, 
                              levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$HeartDiseaseorAttack <- factor(copy_diab.df$HeartDiseaseorAttack,
                                            levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$PhysActivity <- factor(copy_diab.df$PhysActivity, 
                                    levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$Veggies <- factor(copy_diab.df$Veggies, levels = c(0, 1),
                               labels = c("No", "Yes"))

copy_diab.df$HvyAlcoholConsump <- factor(copy_diab.df$HvyAlcoholConsump, 
                                         levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$AnyHealthcare <- factor(copy_diab.df$AnyHealthcare, 
                                     levels = c(0, 1), labels = c("No", "Yes"))

copy_diab.df$NoDocbcCost <- factor(copy_diab.df$NoDocbcCost, 
                                   levels = c(0, 1), labels = c("No", "Yes"))
```


```{r}
str(copy_diab.df)
View(copy_diab.df)
```


```{r distributions}
Diagnosis_distribution <- table(copy_diab.df$DiabOutcome)
print("Diabetes Diagnosis")
print(Diagnosis_distribution)

# Distribution of HighBP
HighBP_distribution <- table(copy_diab.df$HighBP)
print("High Blood Pressure")
print(HighBP_distribution)

# Distribution of HighChol
HighChol_distribution <- table(copy_diab.df$HighChol)
print("High Cholesterol")
print(HighChol_distribution)

# Distribution of CholCheck
CholCheck_distribution <- table(copy_diab.df$CholCheck)
print("Cholesterol Check")
print(CholCheck_distribution)

# Distribution of BMI
BMI_distribution <- table(copy_diab.df$BMI)
print("BMI Distribution")
print(BMI_distribution)

# Distribution of Smoker
Smoker_distribution <- table(copy_diab.df$Smoker)
print("Smoke")
print(Smoker_distribution)

# Distribution of HeartDiseaseorAttack
HeartDiseaseorAttack_distribution <- table(copy_diab.df$HeartDiseaseorAttack)
print("Heart Disease - CHD or MI")
print(HeartDiseaseorAttack_distribution)

# Distribution of PhysActivity
PhysActivity_distribution <- table(copy_diab.df$PhysActivity)
print("Physical Activity:")
print(PhysActivity_distribution)

# Distribution of Veggies
Veggies_distribution <- table(copy_diab.df$Veggies)
print("Vegetables Consumption")
print(Veggies_distribution)

# Distribution of HvyAlcoholConsump
HvyAlcoholConsump_distribution <- table(copy_diab.df$HvyAlcoholConsump)
print("Heavy Alcohol Consumption")
print(HvyAlcoholConsump_distribution)

# Distribution of AnyHealthcare
AnyHealthcare_distribution <- table(copy_diab.df$AnyHealthcare)
print("Health Insurance Coverage")
print(AnyHealthcare_distribution)

# Distribution of NoDocbcCost
NoDocbcCost_distribution <- table(copy_diab.df$NoDocbcCost)
print("No Doctor Visit Due to Cost")
print(NoDocbcCost_distribution)

# Distribution of GenHlth
GenHlth_distribution <- table(copy_diab.df$GenHlth)
print("General Health Status")
print(GenHlth_distribution)

# Distribution of MentHlth
MentHlth_distribution <- table(copy_diab.df$MentHlth)
print("Mental Health Check")
print(MentHlth_distribution)

# Distribution of PhysHlth
PhysHlth_distribution <- table(copy_diab.df$PhysHlth)
print("Physical Health Check")
print(PhysHlth_distribution)

# Distribution of Sex
Sex_distribution <- table(copy_diab.df$Sex)
print("Gender Distribution")
print(Sex_distribution)

# Distribution of Age
Age_distribution <- table(copy_diab.df$Age)
print("Age Distribution")
print(Age_distribution)

# Distribution of Income
Income_distribution <- table(copy_diab.df$Income)
print("Income Distribution")
print(Income_distribution)


```

```{r copy summary}
summary(copy_diab.df)

```

```{r visualization of main variables}

# target variable is "Diabetes Diagnosis" because it gives the diagnosis outcome: No diabetes, Prediabetes, and Diabetes. 

ggplot(copy_diab.df, aes(DiabOutcome)) +
  geom_bar(fill = "blue") +
  labs(title = "Distribution of Diabetes Status")

```
```{r BMI}
ggplot(copy_diab.df, aes(DiabOutcome, BMI, fill = DiabOutcome)) +
  geom_boxplot() +
  labs(title = "BMI vs. Diabetes Status")

ggplot(copy_diab.df, aes(BMI)) +
  geom_bar(fill = "steelblue") +
  labs(title = "BMI Distribution")
```

```{r Age}
ggplot(copy_diab.df, aes(DiabOutcome, Age, fill = DiabOutcome)) +
  geom_boxplot() +
  labs(title = "Age vs. Diabetes Status")

ggplot(copy_diab.df, aes(Age)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Age Distribution")
```


```{r Chol and BP}
ggplot(copy_diab.df, aes(DiabOutcome, HighChol, fill = DiabOutcome)) +
  geom_boxplot() +
  labs(title = "High Cholesterol vs. Diabetes Status")

ggplot(copy_diab.df, aes(HighChol)) +
  geom_bar(fill = "steelblue") +
  labs(title = "High Cholesterol Distribution")

ggplot(copy_diab.df, aes(DiabOutcome, HighBP, fill = DiabOutcome)) +
  geom_boxplot() +
  labs(title = "High Blood Pressure vs. Diabetes Status")

ggplot(copy_diab.df, aes(Sex)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Gender Distribution")
```

```{r}
ggplot(copy_diab.df, aes(DiabOutcome, PhysActivity, fill = DiabOutcome)) +
  geom_boxplot() +
  labs(title = "Age vs. Diabetes Status")

ggplot(copy_diab.df, aes(Age)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Age Distribution")
```



```{r numeric columns visualizations}


numeric_cols <- sapply(copy_diab.df, is.numeric)
copy_diab_numeric <- copy_diab.df[, numeric_cols]

# Convert the numeric data to long format
copy_diab_long <- pivot_longer(copy_diab_numeric,
                               cols = everything(),
                               names_to = "variable",
                               values_to = "value")

# Create the histograms using ggplot2
ggplot(copy_diab_long, aes(x = value)) +
  geom_histogram(bins = 20, fill = "deepskyblue2", color = "black") +
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Histograms of BMI, Mental Health, and Physical Health in The Diabetes Indicator Dataset",
       x = "Value",
       y = "Frequency") +
  theme_bw() +
  theme(strip.text = element_text(size = 8),
        plot.title = element_text(size = 12, hjust = 0.5))
```

```{r}
# Scatter plots (all pairs)
pairs(copy_diab.df[, c("BMI", "MentHlth", "PhysHlth")],
      main = "Scatter Plot Matrix of BMI, MentHlth, PhysHlth")

# Correlation matrix
correlation_matrix <- cor(copy_diab.df[, c("BMI", "MentHlth", "PhysHlth")], use = "complete.obs")
print("Correlation Matrix")
print(correlation_matrix)

# Covariance matrix
covariance_matrix <- cov(copy_diab.df[, c("BMI", "MentHlth", "PhysHlth")], use = "complete.obs")
print("Covariance Matrix")
print(covariance_matrix)
```
# Modelling and Prediction
Categories:

1. Socioeconomic factors
2. Lifestyle factors
3. Health/Medical factors

Models:

1. Logistic regression
2. Random forest
3. Gradient boosting

Research question and goal:

Both revolve around testing the different models for useful predicting diabetes AND finding out which models are useful (IMPORTANT) in successfully predicting diabetes; I want to know the influential predictors, the once that really hold higher meaning.

```{r training}

# using my copy_diab.df and focusing on the diabetes outcome since it is the overall purpose

set.seed(123)

sp_index <- createDataPartition(copy_diab.df$DiabOutcome, p = 0.7, list = FALSE)
train <- copy_diab.df[sp_index, ]
test <- copy_diab.df[-sp_index, ]

```


# Socio economic factors

```{r logrse}

socio_logm <- glm(DiabOutcome ~ Age + Sex + Income + GenHlth + AnyHealthcare + NoDocbcCost,
                  data = train, 
                  family = binomial)

summary(socio_logm)
```

```{r rfse}

socio_rf <- randomForest(DiabOutcome ~ Age + Sex + Income + GenHlth + AnyHealthcare + NoDocbcCost, data = train,
                         importance = TRUE)
varImpPlot(socio_rf)
```

```{r}
plot(socio_rf)
```


```{r gbmse}
train$DiabNumer <- ifelse(train$DiabOutcome == "Diabetes", 1, 0)
test$DiabNumer  <- ifelse(test$DiabOutcome == "Diabetes", 1, 0)
```

```{r}
table(train$DiabNumer)
str(train$DiabNumer) 
```

```{r gbmse}
socio_gbm <- gbm(DiabNumer ~ Age + Sex + Income + GenHlth + AnyHealthcare + NoDocbcCost, 
                 data = train,
                 distribution = "bernoulli",
                 n.trees = 1000,
                 interaction.depth = 4,
                 shrinkage = 0.01,
                 cv.folds = 5,
                 n.cores = NULL,
                 verbose = FALSE)

bestnotreese <- gbm.perf(socio_gbm, method = "cv")
```


```{r predictevaluatemodels}
# LOGR

sociologm_probs <- predict(socio_logm, test, type = "response")
sociologm_pred <- ifelse(sociologm_probs > 0.5, "Diabetes", "No Diabetes")
sociologm_pred <- factor(sociologm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(sociologm_pred, test$DiabOutcome)
```

```{r predictevaluatemodels}
# RF

sociorf_pred <- predict(socio_rf, test)
confusionMatrix(sociorf_pred, test$DiabOutcome)
```

```{r predictevaluatemodels}
# GBM

sociogbm_probs <- predict(socio_gbm, test, n.trees = bestnotreese, type = "response")
sociogbm_pred <- ifelse(sociogbm_probs > 0.5, "Diabetes", "No Diabetes")
sociogbm_pred <- factor(sociogbm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(sociogbm_pred, test$DiabOutcome)

```

```{r}

# Logistic Regression metrics
logm_cm <- confusionMatrix(sociologm_pred, test$DiabOutcome)
logm_acc <- logm_cm$overall['Accuracy']
logm_sens <- logm_cm$byClass['Sensitivity']
logm_spec <- logm_cm$byClass['Specificity']
logm_prec <- logm_cm$byClass['Pos Pred Value']
logm_f1 <- 2 * (logm_prec * logm_sens) / (logm_prec + logm_sens)

# Random Forest metrics
rf_cm <- confusionMatrix(sociorf_pred, test$DiabOutcome)
rf_acc <- rf_cm$overall['Accuracy']
rf_sens <- rf_cm$byClass['Sensitivity']
rf_spec <- rf_cm$byClass['Specificity']
rf_prec <- rf_cm$byClass['Pos Pred Value']
rf_f1 <- 2 * (rf_prec * rf_sens) / (rf_prec + rf_sens)

# GBM metrics
gbm_cm <- confusionMatrix(sociogbm_pred, test$DiabOutcome)
gbm_acc <- gbm_cm$overall['Accuracy']
gbm_sens <- gbm_cm$byClass['Sensitivity']
gbm_spec <- gbm_cm$byClass['Specificity']
gbm_prec <- gbm_cm$byClass['Pos Pred Value']
gbm_f1 <- 2 * (gbm_prec * gbm_sens) / (gbm_prec + gbm_sens)

```

```{r}
modelsrlsts <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "GBM"),
  Category = "Socioeconomic",
  Accuracy = c(logm_acc, rf_acc, gbm_acc),
  Precision = c(logm_prec, rf_prec, gbm_prec),
  Recall = c(logm_sens, rf_sens, gbm_sens),   # aka Sensitivity
  Specificity = c(logm_spec, rf_spec, gbm_spec),
  F1_Score = c(logm_f1, rf_f1, gbm_f1)
)

print(modelsrlsts)

```



```{r}
write.csv(modelsrlsts, "modelsrlsts.cvs", row.names = FALSE)

```

Dependent variable:
•	Diabetes Diagnosis (3 classes: No Diabetes, Prediabetes, and Diabetes)
Independent Variables: 
•	Sensitive demographic: Sex, Age, and Income
•	Lifestyle activities: Smoker, PhysActivity, Veggies, and HvyAlcoholConsump
•	Health Status: HighBP, HighChol, Stroke, Heart DiseaseorAttak, CholCheck, BMI, GenHlth, MentHlth, and PhysHlth
•	Healthcare access: AnyHealthCare, NoDocbcCost


# Lifestyle factors

```{r logls}

lifes_logm <- glm(DiabOutcome ~ Smoker + PhysActivity + Veggies + HvyAlcoholConsump, 
                  data = train, 
                  family = binomial)

summary(lifes_logm)

```

```{r rfls}
lifes_rf <- randomForest(DiabOutcome ~ Smoker + PhysActivity + Veggies + HvyAlcoholConsump,
                         data = train,
                         importance = TRUE)
varImpPlot(lifes_rf)

```
```{r}
plot(lifes_rf)
```

```{r gbmls}
lifes_gbm <- gbm(DiabNumer ~ Smoker + PhysActivity + Veggies + HvyAlcoholConsump, 
                 data = train,
                 distribution = "bernoulli",
                 n.trees = 1000,
                 interaction.depth = 4,
                 shrinkage = 0.01,
                 cv.folds = 5,
                 n.cores = NULL,
                 verbose = FALSE)

bestnotreels <- gbm.perf(lifes_gbm, method = "cv")
```


```{r lifestyleprediction}
# LOGR  

lifeslogm_probs <- predict(lifes_logm, test, type = "response")
lifeslogm_pred <- ifelse(lifeslogm_probs > 0.3, "Diabetes", "No Diabetes")
lifeslogm_pred <- factor(lifeslogm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(lifeslogm_pred, test$DiabOutcome)
```


```{r lifestyleprediction}
# RF

lifesrf_pred <- predict(lifes_rf, test)
confusionMatrix(lifesrf_pred, test$DiabOutcome)

```


```{r lifestyleprediction}
# GBM

lifesgbm_probs <- predict(lifes_gbm, test, n.trees = bestnotreels, type = "response")
lifesgbm_pred <- ifelse(lifesgbm_probs > 0.3, "Diabetes", "No Diabetes")
lifesgbm_pred <- factor(lifesgbm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(lifesgbm_pred, test$DiabOutcome)
```


```{r life}

# Logistic Regression metrics
lifeslog_cm <- confusionMatrix(lifeslogm_pred, test$DiabOutcome)
lslogm_acc <- lifeslog_cm$overall['Accuracy']
lslogm_sens <- lifeslog_cm$byClass['Sensitivity']
lslogm_spec <- lifeslog_cm$byClass['Specificity']
lslogm_prec <- lifeslog_cm$byClass['Pos Pred Value']
lslogm_f1 <- 2 * (lslogm_prec * lslogm_sens) / (lslogm_prec + lslogm_sens)

# Random Forest metrics
lifesrf_cm <- confusionMatrix(lifesrf_pred, test$DiabOutcome)
lsrf_acc <- lifesrf_cm$overall['Accuracy']
lsrf_sens <- lifesrf_cm$byClass['Sensitivity']
lsrf_spec <- lifesrf_cm$byClass['Specificity']
lsrf_prec <- lifesrf_cm$byClass['Pos Pred Value']
lsrf_f1 <- 2 * (lsrf_prec * lsrf_sens) / (lsrf_prec + lsrf_sens)

# GBM metrics
lifesgbm_cm <- confusionMatrix(lifesgbm_pred, test$DiabOutcome)
lsgbm_acc <- lifesgbm_cm$overall['Accuracy']
lsgbm_sens <- lifesgbm_cm$byClass['Sensitivity']
lsgbm_spec <- lifesgbm_cm$byClass['Specificity']
lsgbm_prec <- lifesgbm_cm$byClass['Pos Pred Value']
lsgbm_f1 <- 2 * (lsgbm_prec * lsgbm_sens) / (lsgbm_prec + lsgbm_sens)

#

lsmodelsrlsts <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "GBM"),
  Category = "Lifestyle",
  Accuracy = c(lslogm_acc, lsrf_acc, lsgbm_acc),
  Precision = c(lslogm_prec, lsrf_prec, lsgbm_prec),
  Recall = c(lslogm_sens, lsrf_sens, lsgbm_sens),   # aka Sensitivity
  Specificity = c(lslogm_spec, lsrf_spec, lsgbm_spec),
  F1_Score = c(lslogm_f1, lsrf_f1, lsgbm_f1)
)

print(lsmodelsrlsts)

```




# Medical factors

```{r loghf}

medhlth_logm <- glm(DiabOutcome ~ HighBP + HighChol + CholCheck + BMI + HeartDiseaseorAttack + MentHlth + PhysHlth,
                  data = train, 
                  family = binomial)

summary(medhlth_logm)

```

```{r rfhf}
medhlth_rf <- randomForest(DiabOutcome ~ HighBP + HighChol + CholCheck + BMI + HeartDiseaseorAttack + MentHlth + PhysHlth,
                         data = train,
                         importance = TRUE)
varImpPlot(medhlth_rf)

```
```{r}
plot(medhlth_rf)
```

```{r gbmhf}
medhlth_gbm <- gbm(DiabNumer ~ HighBP + HighChol + CholCheck + BMI + HeartDiseaseorAttack + MentHlth + PhysHlth, 
                 data = train,
                 distribution = "bernoulli",
                 n.trees = 1000,
                 interaction.depth = 4,
                 shrinkage = 0.01,
                 cv.folds = 5,
                 n.cores = NULL,
                 verbose = FALSE)

bestnotreemh <- gbm.perf(medhlth_gbm, method = "cv")
```


```{r lifestyleprediction}
# LOGR  

mhlogm_probs <- predict(medhlth_logm, test, type = "response")
mhlogm_pred <- ifelse(mhlogm_probs > 0.3, "Diabetes", "No Diabetes")
mhlogm_pred <- factor(mhlogm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(mhlogm_pred, test$DiabOutcome)
```


```{r lifestyleprediction}
# RF

mhrf_pred <- predict(medhlth_rf, test)
confusionMatrix(mhrf_pred, test$DiabOutcome)

```


```{r lifestyleprediction}
# GBM

mhgbm_probs <- predict(medhlth_gbm, test, n.trees = bestnotreemh, type = "response")
mhgbm_pred <- ifelse(mhgbm_probs > 0.3, "Diabetes", "No Diabetes")
mhgbm_pred <- factor(mhgbm_pred, levels = c("No Diabetes", "Diabetes"))

confusionMatrix(mhgbm_pred, test$DiabOutcome)
```



```{r life}

# Logistic Regression metrics
mhlog_cm <- confusionMatrix(mhlogm_pred, test$DiabOutcome)
mhlogm_acc <- mhlog_cm$overall['Accuracy']
mhlogm_sens <- mhlog_cm$byClass['Sensitivity']
mhlogm_spec <- mhlog_cm$byClass['Specificity']
mhlogm_prec <- mhlog_cm$byClass['Pos Pred Value']
mhlogm_f1 <- 2 * (mhlogm_prec * mhlogm_sens) / (mhlogm_prec + mhlogm_sens)

# Random Forest metrics
mhrf_cm <- confusionMatrix(mhrf_pred, test$DiabOutcome)
mhrf_acc <- mhrf_cm$overall['Accuracy']
mhrf_sens <- mhrf_cm$byClass['Sensitivity']
mhrf_spec <- mhrf_cm$byClass['Specificity']
mhrf_prec <- mhrf_cm$byClass['Pos Pred Value']
mhrf_f1 <- 2 * (mhrf_prec * mhrf_sens) / (mhrf_prec + mhrf_sens)

# GBM metrics
mhgbm_cm <- confusionMatrix(mhgbm_pred, test$DiabOutcome)
mhgbm_acc <- mhgbm_cm$overall['Accuracy']
mhgbm_sens <- mhgbm_cm$byClass['Sensitivity']
mhgbm_spec <- mhgbm_cm$byClass['Specificity']
mhgbm_prec <- mhgbm_cm$byClass['Pos Pred Value']
mhgbm_f1 <- 2 * (mhgbm_prec * mhgbm_sens) / (mhgbm_prec + mhgbm_sens)

#

mhmodelsrlsts <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "GBM"),
  Category = "MedHealth",
  Accuracy = c(mhlogm_acc, mhrf_acc, mhgbm_acc),
  Precision = c(mhlogm_prec, mhrf_prec, mhgbm_prec),
  Recall = c(mhlogm_sens, mhrf_sens, mhgbm_sens),   # aka Sensitivity
  Specificity = c(mhlogm_spec, mhrf_spec, mhgbm_spec),
  F1_Score = c(mhlogm_f1, mhrf_f1, mhgbm_f1)
)

print(mhmodelsrlsts)

```



## propensity score

```{r}
treatment <- "Smoker" # 1 = Yes, 0 = No
outcome <- "DiabOutcome"

```

```{r}
ps_model <- glm(Smoker ~ Age + Sex + Income + GenHlth + PhysActivity + BMI + HvyAlcoholConsump,
                data = copy_diab.df,
                family = binomial)

copy_diab.df$pscore <- predict(ps_model, type = "response")

```

```{r}
library(MatchIt)
match_it <- matchit(Smoker ~ Age + Sex + Income + GenHlth + PhysActivity + BMI + HvyAlcoholConsump,
                    data = copy_diab.df, distance = "glm", method = "nearest", 
                    m.order = "largest", replace = FALSE)

matched_data <- match.data(match_it)

```

```{r}
summary(matched_data)
```

```{r}
summary(glm(DiabOutcome ~ Smoker, data = matched_data, family = binomial))
```

```{r}
treatment1 <- "HeartDisease" # 1 = Yes, 0 = No
outcome <- "DiabOutcome"

```

```{r}
ps_model1 <- glm(HeartDiseaseorAttack ~ Age + HighBP + HighChol + GenHlth + PhysActivity + BMI + HvyAlcoholConsump,
                data = copy_diab.df,
                family = binomial)

copy_diab.df$pscore <- predict(ps_model1, type = "response")

```

```{r}
#PSA

match_it1 <- matchit(HeartDiseaseorAttack ~ Age + HighBP + HighChol + GenHlth + PhysActivity + BMI + HvyAlcoholConsump,
                    data = copy_diab.df, distance = "glm", method = "nearest", 
                    m.order = "largest", replace = FALSE)

matched_data1 <- match.data(match_it1)

```

```{r}
summary(matched_data1)
```

```{r}
summary(glm(DiabOutcome ~ HeartDiseaseorAttack, data = matched_data1, family = binomial))
```

```{r}
treatment2 <- "NoDocbcCost" # 1 = Yes, 0 = No
outcome2 <- "DiabOutcome"

```

```{r}
ps_model2 <- glm(NoDocbcCost ~ Age + PhysActivity + AnyHealthcare + BMI + Sex + MentHlth,
                data = copy_diab.df,
                family = binomial)

copy_diab.df$pscore <- predict(ps_model2, type = "response")

```

```{r}

match_it2 <- matchit(NoDocbcCost ~ Age + PhysActivity + AnyHealthcare + BMI + Sex + MentHlth,
                    data = copy_diab.df, distance = "glm", method = "nearest", 
                    m.order = "largest", replace = FALSE)

matched_data2 <- match.data(match_it2)

```

```{r}
summary(matched_data2)
```

```{r}
summary(glm(DiabOutcome ~ NoDocbcCost, data = matched_data2, family = binomial))
```
## PSA propensity



